---
title: "EDA"
author: "Rohan Patel"
date: "2023-11-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

if (!requireNamespace("ggplot2", quietly = TRUE)) {
  install.packages("ggplot2")
}
library(ggplot2)

if (!requireNamespace("psych", quietly = TRUE)) {
  install.packages("psych")
}
library(psych)

if (!requireNamespace("reshape2", quietly = TRUE)) {
  install.packages("reshape2")
}
library(reshape2)
```


# Importing the Data
```{r}
setwd("..")
data <- read_csv('Data/cleaned_data.csv')
head(data)
```


# Exploring the Data
```{r}
pairs.panels(data)
```


```{r}
ggplot(data, aes(x = as.factor(Diagnosis), fill = as.factor(Diagnosis))) +
  geom_bar() +
  labs("Countplot of Diagnosis", x = "Diagnosis") +
  scale_fill_discrete(name = "Diagnosis")
```
Majority of patients have COPD or are parts of Healthy control groups, data is imbalanced


```{r}
ggplot(data, aes(x = as.factor(Diagnosis), fill = as.factor(Gender))) +
  geom_bar(position = "dodge") +
  labs("Countplot of Diagnosis with Hue Gender", x = "Diagnosis") +
  scale_fill_discrete(name = "Gender")
```
More Female paitents have Asthma and Infections then males


```{r}
ggplot(data, aes(x = as.factor(Gender), fill = as.factor(Gender))) +
  geom_bar() +
  labs("Countplot of Gender", x = "Gender") +
  scale_fill_discrete(name = "Gender")
```
As always majority of paitents are males



```{r}
ggplot(data, aes(x = as.factor(Smoking), fill = as.factor(Smoking))) +
  geom_bar() +
  labs("Countplot of Smoking", x = "Smoking") +
  scale_fill_discrete(name = "Smoking")
```
Most paitients are ex-smokers and non smokers, a few active smokers.


```{r}
plot1 = ggplot(data, aes(x = Age)) +
  geom_histogram(binwidth = 5, fill = "lightblue", color = "black", alpha = 0.7) +
  ggtitle("Distribution of Age") +
  xlab("Age")

plot2 = ggplot(data, aes(x = Age)) +
  geom_density(fill = "lightblue", color = "black", alpha = 0.7) +
  ggtitle("Distribution of Age") +
  xlab("Age")

grid.arrange(plot1, plot2, ncol = 1)
```
All paitents are adults, ranging from young adults in the 20's to senior citizens upto their late seventies, very few paitents in their eighties and beyond



```{r}
ggplot(data, aes(x = as.factor(Smoking), fill = as.factor(Gender))) +
  geom_bar(position = "dodge") +
  labs("Countplot of Smoking with Hue Gender", x = "Smoking") +
  scale_fill_discrete(name = "Gender")
```



```{r}
# Disease by non smokers
ggplot(data, aes(x = as.factor(Smoking), fill = as.factor(Diagnosis))) +
  geom_bar(position = "dodge") +
  labs("Countplot of Smoking with Hue Diagnosis", x = "Smoking") +
  scale_fill_discrete(name = "Diagnosis")
```
Non-Smokers:
most are healthy, rest have infections and asthma, very few with COPD

Ex-Smokers:
Majority have COPD.

Active Smokers:
Not cases with infections, most are healthy with few having asthama and COPD


```{r}
# Disease by non smokers
ggplot(data, aes(x = Diagnosis, y = Age, group = Diagnosis)) +
  geom_boxplot(fill = "lightblue", color = "black", alpha = 0.7) +
  ggtitle("Boxplot of Age by Diagnosis") +
  xlab("Diagnosis") +
  ylab("Age")
```
COPD:
Seen mostly in middle aged to senior citizens

Asthama:
Obsevered across almost all age groups

Infections:
Not seen in older paients beyond 50-55

Healthy Groups:
Oldest paients are upto the age of 60/65


```{r}
ggplot(data, aes(x = `Imaginary Part: Min`)) +
  geom_density(fill = "lightblue", color = "black", alpha = 0.7) +
  ggtitle("Distribution Plot of Imaginary Part: Min") +
  xlab("Imaginary Part: Min")
```



```{r}
ggplot(data, aes(x = `Imaginary Part: Avg`)) +
  geom_density(fill = "lightblue", color = "black", alpha = 0.7) +
  ggtitle("Distribution Plot of Imaginary Part: Avg") +
  xlab("Imaginary Part: Avg")
```



```{r}
ggplot(data, aes(x = `Real Part: Min`)) +
  geom_density(fill = "lightblue", color = "black", alpha = 0.7) +
  ggtitle("Distribution Plot of Real Part: Min") +
  xlab("Real Part: Min")
```



```{r}
ggplot(data, aes(x = `Real Part: Avg`)) +
  geom_density(fill = "lightblue", color = "black", alpha = 0.7) +
  ggtitle("Distribution Plot of Real Part: Avg") +
  xlab("Real Part: Avg")
```
Imaginary part and real part data has a wide distribution with outlier values



```{r}
correlation_long <- melt(cor(data))
ggplot(correlation_long, aes(Var1, Var2, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

